Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-04-01 doc/paper.Rmd
revise intro slightly

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017].
These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
All of these factors have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
Plants, unlike animals, are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory resulting from fragmentation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].

Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
First, most studies of plants in fragments have focused on a single life-history stage or process ***(CITATIONS)*** <!--# citations in proposal -->, making it challenging to draw broader demographic conclusions.
Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
These data are critical not only because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].

We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
<!--# add a sentence about "famous" amazonian droughts --> Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
We addressed the following questions:

@@

<!-- Rainfall data should prob be a figure in main text, since the paper deals with rainfall and seasonality. I’d think of citing the (eventual) BDFFP rain package -->

@@

-   @horvitz2002 Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
    Survival of large stages was not generally affected by any of the factors.
    Relative growth rates of all stages were negatively affected by plant size; larger plants grew more slowly than smaller plants.
    Additionally, growth rate of juveniles was negatively affected by local competition, whereas growth rate of reproductives was positively affected by reproduction.
    Next year's inflorescence production was positively affected by plant size for both pre-reproductive and reproductive stages.
    For reproductives, it was additionally positively affected by current fruit production.

>>>>

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017].
These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
<!--# Additionally, extreme wet episodes are becoming as severe and frequent as droughts in parts of the Amazon, and can also have negative effects on plants [@estebanOtherSideDroughts2021].  -->All of these factors have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable.
Plants, unlike animals, are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory resulting from fragmentation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017].
Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@bettsExtinctionFiltersMediate2019].

Whether plant populations in tropical forest fragments are particularly susceptible to drought remains unclear for three primary reasons.
First, most studies of plants in fragments have focused on a single life-history stage or process [@brunaHabitatFragmentationPlant2009; @ehrlenAdvancingEnvironmentallyExplicit2016], making it challenging to draw broader demographic conclusions.
Similarly, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments [@lauranceEffectsStrongDrought2001].
Finally, the multi-year data needed to test population-level hypotheses about climate change and fragmentation are scant, especially for tropical systems [@salguero-gomezCOMPADREPlantMatrix2015; @croneHowPlantEcologists2011].
These data are critical not only because they allow for capturing variation in climatic conditions and the resulting demographic responses [@morrisQuantitativeConservationBiology2002; @tellerLinkingDemographyDrivers2016].
They are also essential because while some demographic effects of fragmentation or drought can be detected immediately, others may take years to manifest [@gagnonGrowthUnderstoryHerb2011].
Indeed, lagged responses of demographic vital rates to climate may in fact be the rule rather than the exception [@eversLaggedDormantSeason2021; @andereggPervasiveDroughtLegacies2015; @schwalmGlobalPatternsDrought2017; @kannenbergGhostsHowDrought2020].

We used a decade of demographic and climatic data to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae) in an experimentally fragmented landscape in the Central Amazon.
This time series includes "famous" Amazonian droughts in 1997 [@mcphadenChildProdigy1997981999] and 2005[@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008] that were called "once in a century" climate events.
Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***]{.ul}.
We addressed the following questions:

@@

@@

-   @horvitz2002 (*Calathea ovandensis*) Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
    Survival of large stages was not generally affected by any of the factors.
    Relative growth rates of all stages were negatively affected by plant size; larger plants grew more slowly than smaller plants.
    Additionally, growth rate of juveniles was negatively affected by local competition, whereas growth rate of reproductives was positively affected by reproduction.
    Next year's inflorescence production was positively affected by plant size for both pre-reproductive and reproductive stages.
    For reproductives, it was additionally positively affected by current fruit production.
